Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-01 demystify.tex
updates

A LoR writer should be someone who knows you well, so that they can provide a strong-supporting and personalized letter for you.  
LoR writers should also have the credibility to convince adcom members. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability).
LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school).

@@

However, let's face it.  Most of us don't have an opportunity to interact or work with well-known researchers.  
So what to do? It is perfectly fine to have a letter from people who know you well enough to talk about \emph{your research experience and capabilities}. These include profs., researchers, postdocs who have mentored and worked with you. In fact, a personal, customized, and strong letter from even a senior PhD grad student who mentors and knows you well is much better than a generic letter from a well-known superstar who does not know you well.

@@

This does not mean the professor is not good or does not care about you, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
 Many international students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter). Experienced reviewers can easily spot  (because the student often has little experience in writing LoRs) and  \red{red flag} such letters.

@@

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. For example, the following list several LOR invitation emails I received from various universities in the past few years:

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work at such places, then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LORs (\autoref{sec:lor}).

\begin{commentbox}[Hung:]
  The reputation of VinAI has been increasing steadily over the past few years; many of my colleagues heard about VinAI.
\end{commentbox}

On the other hand, if non-research work (e.g., software development) \emph{do not} carry much weight for PhD (they can help for MS application though). Similarly, LoR from your supervisor for non-research experience will not count much.
So do not spend much time talking about non-research-related job in your SOP.

>>>>

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also have the credibility to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability). LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\paragraph{LoR from Famous People}  Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school). The reason is simple: adcom members trust people that they know and respect. If a well-known researcher enthusiastically recommends you, then the adcom members will likely trust that recommendation.

@@

However, let's face it.  Most students don't have an opportunity to interact or work with well-known researchers.  
So what to do? It is perfectly fine to have a letter from people who know you well enough to talk about \emph{your research experience and capabilities}. These include profs., researchers, postdocs who have mentored and worked with you. In fact, a personal, customized, and strong letter from even a senior PhD grad student who mentors and knows you well is much better than a generic letter from a well-known superstar who does not know you well.

@@

\emph{This does not mean the professor is not good or does not care about you}, but they just do not know you well enough to write a strong letter.
So it might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
Many students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

Adcom members are just too experienced (having reviews hundreds of those letters) to spot and suspect such letters.  There is a very difference between a letter written by an experienced professor and a student.  For example, they can provide convincing and concrete examples based on their experience and compare you to their own students, and of course the writing style of a professor is very different from an undergraduate student.  Worse yet, if we suspect that the student wrote the letter, it is a \red{red flag} as we will question both the student's integrity and the letter writer's credibility.

@@

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. The following lists several LOR invitation emails I received from various universities in the past few years:

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Note that the emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer is not research experience, though it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about non-research job in your SOP.

Note that adcom reviewers might not be familiar with all research labs, especially those outside the US For example, while VinAI is  well-known in Vietnam and potentially in Asia, it might not be very well-known in the US. So you or your LoR writers should explicitly say something about them in your statement.  In general, if you did some good research work, then you should mention that in your SOP and ask your supervisor to write about it in their LOR (\autoref{sec:lor}).

% \begin{commentbox}[Hung:]
%   The reputation of VinAI has been increasing steadily over the past few years; many of my colleagues heard about VinAI.
% \end{commentbox}
